---
title: "Analyzing Bike Rides"
author: "FT"
date: "Sunday, January 25, 2015"
output: html_document
---



```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(stringr)
library(lubridate)
library(ggplot2)
library(plyr)
library(grid)
library(gridExtra)
library(stargazer)
raw <- read.csv("out.txt")
raw$V1 <- as.character(raw$X.200)
```

I pulled last year's data from the "CallaBike"-SOAP API.

```{r,}
length(grep("StartTime", raw$V1,)) 
```

So it looks like  did 403 trips using a Call a Bike in 2014.
Next we need some data cleaning.
```{r, echo=FALSE}
trips <- data.frame(cbind(raw[grep("StartTime", raw$V1,),]$V1, raw[grep("EndTime", raw$V1,),]$V1) )

trips$start = str_trim(str_split_fixed(trips$X1, "=", 2)[,2])
trips$end = str_trim(str_split_fixed(trips$X2, "=", 2)[,2])

#as.POSIXlt
trips$startT <- as.POSIXct(trips$start, format="%Y-%m-%d %H:%M:%S", tz = "CET")
trips$endT <- as.POSIXct(trips$end, format="%Y-%m-%d %H:%M:%S")

## summertime!
#interv <- as.interval(as.Date("30.03.2014", "%d.%m.%Y"), as.Date("26.10.2014", "%d.%m.%Y"))
daylight <- as.interval(ymd_hm("2014-03-30 00:03"), ymd_hm("2014-10-26 24:23"))

#trips$daylight <- ifelse(trips$startT %within% daylight, 1,0)

trips[trips$startT %within% daylight,]$startT <- trips[trips$startT %within% daylight,]$startT +60*60
trips[trips$startT %within% daylight,]$endT <- trips[trips$startT %within% daylight,]$endT +60*60

trips$driveTime <- (trips$endT - trips$startT )/ 60 
trips$mon<- format(trips$startT, "%m %b")
trips$wday <- format(trips$startT, "%w %a")
trips$hour <- hour(trips$startT)
## morning trips!
#table(trips$hour)
trips$morning <- ifelse(trips$hour %in% c(7,8),1,0)
trips$evening <- ifelse(trips$hour %in% c(15,16,17,18,19,20),1,0)
## remove
print("MAX drive time:")
max(trips$driveTime)  #Time difference of 9571.433 mins

trips <- trips[trips$driveTime <60,]
trips$dt <- as.numeric(trips$driveTime)

trips$timeZ <- "rest"
trips$timeZ <-ifelse(trips$evening==1, "evening", trips$timeZ)
trips$timeZ <-ifelse(trips$morning==1, "morning", trips$timeZ)
```
remove an unreasonable high drive time. Next we will get an initial glimpse at the data.

```{r, echo=FALSE, message=FALSE}
p1=ggplot(trips, aes(hour)) + geom_bar() + ggtitle("# trips hour") 

p2=ggplot(trips, aes(wday)) + geom_bar() + ggtitle("# trips weekday") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

p3=ggplot(trips, aes(mon)) + geom_bar() + ggtitle("# trips month")+ theme(axis.text.x = element_text(angle = 90, hjust = 1))
grid.arrange(p1, p2, p3, nrow=1)
```

We see a clear pattern of high usage during the week and at commuting hours.

```{r, echo=FALSE, message=FALSE}
ggplot(trips, aes(timeZ, dt  )) + geom_boxplot() + xlab("") + ylab("commute time")


```

Commute times are basically the same for the morning/evening trips.

```{r, echo=FALSE, message=FALSE}
ggplot(trips, aes(mon, dt, group=mon )) + geom_boxplot()
```

Commute times did not change over the year. (some ups and downs -but it's basically stable.)

According to google maps the distance is 4.9km. Which makes a total of
```{r, echo=FALSE, message=FALSE}
nrow(trips[trips$evening==1 || trips$morning==1,]) *4.9
```
KM in 2014.
at an average speed of:

```{r, echo=FALSE, message=FALSE}
 4.9/ mean(trips[trips$evening==1 || trips$morning==1,]$dt) *60
```

```{r, echo=FALSE, message=FALSE}
trips$date <- as.Date(trips$startT)
working <- trips[trips$morning ==1,]
working <- merge(working, trips[trips$evening ==1,], by=c("date"))
working <- working[,c("startT.x", "endT.x", "dt.x", "startT.y", "endT.y", "dt.y")]

working$wt <- working$startT.y -working$endT.x
working$wt <- as.numeric(working$wt) ## in hours
working$mon<- format(working$startT.x, "%m %b")
working$wday <- format(working$startT.x, "%w %a")
```
As this data indicates starting and ending of commutes, we can calculate the time spend at work.

```{r, echo=FALSE, message=FALSE}
ggplot(working, aes(x=wt, group=wday, fill=wday)) + geom_density(alpha=.3) + xlab("time spend at work in hours")
mean(working$wt)
```
Looks like an easy 50hours week (on average). With most the time being at work from monday till tuesday.

```{r, echo=FALSE, message=FALSE}
ggplot(working, aes(y=wt, x=mon)) +geom_boxplot()  + ylab("time spend at work in hours") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
lm <- lm(wt~wday+mon+dt.x+dt.y, data=working)
lm1 <- lm(wt~wday+mon, data=working)
lm3 <- lm(dt.y~wday+mon+wt, data=working)
```

And finally some regression model ;)

```{r,results='asis'}
stargazer(lm1, lm,  type = "html", title="Explain time spend at work.")
```

why is time spend at work negatively correlated with drive time?

```{r,results='asis'}
lm3 <- lm(dt.y~wt, data=working)
lm4 <- lm(dt.y~wt+wday, data=working)
lm5 <- lm(dt.y~wt+wday +mon, data=working)

stargazer(lm3,lm4,lm5,  type = "html", title="Explain commute time back from work.")
```
